Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-20 submission/manuscript.Rmd
First pass at Materials and Methods

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model. Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories. Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (P>`r min_p_compare_metag`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@Sze2017; @Baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=**r length(metai$sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf$time)** - **r max(metaf$time)** days. Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@Sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@Sze2017; @Baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@Venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations. These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).

**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @Sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@Westcott2017].

**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used. Briefly, the mothur shared file and metadata was converted into a biom formatted table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.

**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=**r length(metai$sample)**) containing a total of **r length(geof_samples$X1)** individuals (normal n=**r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()**, adenoma n=**r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()**, and carcinoma n=**r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()**) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].

**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@Liaw2002; @Kuhn2017]. This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations. The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [**XXXXXX**]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of genes encoding enzymes involved in SCFA synthesis was tested. Additionally, metagenomic sequencing counts for important genes involved with SCFA production were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without colon tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify the actual SCFA concentration or high/low SCFA concentration based on the median of each SCFA using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.

>>>>

follow_up_time <- read_csv('../data/metadata/follow_up.csv') %>%
 	summarize(q1=quantile(time, prob=0.25), median=median(time), q3=quantile(time, prob=0.75))

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model. Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A). When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories. Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (P>`r min_p_compare_metag`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

@@

**Study design and sampling.** The overall study design and the resulting sequence data have been previously described [@Sze2017; @Baxter2016]. In brief, fecal samples were obtained from `r n_normal` individuals with normal colons, `r n_adenoma` individuals with colonic adenomas, and `r n_cancer` individuals with carcinomas. Of the individuals diagnosed as having adenomas or carcinomas, a subset (N~adenoma~=`r n_follow_adenoma` and N~carcinoma~=`r n_follow_cancer`) were sampled after treatment of the lesion (median=`r follow_up_time$median` days between sampling, IQR=`r follow_up_time$q1` to `r follow_up_time$q3` days). Tumor diagnosis was made by colonoscopic examination and histopathological review of the biopsies [@Baxter2016]. The University of Michigan Institutional Review Board approved the studies that generated the samples and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

**Measuring specific SCFAs.** The measurement of acetate, propionate, isobutyrate, and butyrate used a previously published protocol that used High-Performance Liquid Chromatography (HPLC) [@Venkataraman2016]. Two changes were made to the protocol. First, instead of using fecal samples suspended in DNA Genotek OmniGut tubes, we suspended frozen fecal samples in 1 mL of PBS. Second, instead of using the average weight of fecal sample aliquots to normalize SCFA concentrations, we used the actual weight of the fecal samples. These methodological changes did not affect the range of concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).

**16S rRNA gene sequence data analysis.** Sequence data from Baxter et al. [@Baxter2016] and Sze et al. [@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using using mothur v.1.42 [@Schloss2009]. The sequence generation workflow and processing steps have been previously described [@Schloss2009; @Kozich2013]. In brief, the original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer [@Kozich2013]. The resulting sequence data were assembled into contigs and screened for quality aberrations including chimeras, which were removed based on the use of UCHIME [@Edgar2011]. Sequences were clustered into OTUs at 97% similarity using OptiClust [@Westcott2017] and assigned to the closest possible genus using a naive Bayesian classifier with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16) [@Wang2007]. We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways, KEGG categories, and enzyme commission numbers [@Langille2013].

**Metagenomic DNA sequence analysis.** A subset of the samples from the samples described by Baxter et al. [@Baxter2016] were used to generate metagenomic sequence data (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`). These data were generated by Hannigan et al. [@Hannigan2017] and deposited into the Sequence Read Archive (study SRP108915). Fecal DNA was subjected to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads. The archived sequences were already quality filtered and aligned to the human genome to remove contaminating sequence data. We downloaded the sequences and assembled them into contigs using MEGAHIT [@Li2015], which were used to identify open reading frames (ORFs) using Prodigal [@Hyatt2010]. We determined the abundance of each ORF by mapping the raw reads back to the ORFs using Diamond [@Buchfink2014]. We clustered the ORFs into operational protein families (OPFs) in which the clustered ORFs were more than 40% identical to each other using mmseq2 [@Steinegger2017; @Schloss2008]. We also used mmseq2 to map the ORFs to the KEGG database and clustered the ORFs according to which category the ORFs mapped [@Kanehisa2015].

**Random Forest models.**

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.1) with the tidyverse package (v.1.2.1). To assess differences in SCFA concentrations between individuals normal colons and those with adenomas or carcinomas, we used the Kruskal-Wallis rank sum test. If a test had a P-value below 0.05, we then applied a pairwise Wilcoxon rank sum test with a Benjamini-Hochberg correction for multiple comparisons. To assess differences in SCFA concentrations between individuals samples before and after treatment we used paired Wilcoxon rank sum tests to test for significance. To compare the median AUCROC for the held out data for the model generated using only the SCFAs, we compared the distribution of the data to the expected median of 0.5 using the Wilcoxon rank sum test to test whether the model performed better than would be achieved by randomly assigning the data to each diagnosis. When we compared the Random Forest models generated without and with SCFA data included, we used Wilcoxon rank sum tests to determine whether the models with the SCFA data included did better.

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/.

### 2019-03-20 submission/manuscript.tex
First pass at Materials and Methods

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models (13, 14). We repeated that analysis but
added the concentration of the individual or pooled SCFAs as possible
features to train a model. Models trained using SCFAs to classify
individuals as having adenomas or carcinomas rather than normal colons
had median areas under the receiver operator characteristic curve
(AUROC) that were significantly greater than 0.5
(P\textsubscript{adenoma}\textless{}0.001 and
P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC values
to detect the presence of adenomas or carcinomas were only 0.54 and
0.55, respectively (Figure 2A). When we trained the models with the
SCFAs concentrations and OTU or genus-level relative abundances the
AUROC values were not significantly different from the models trained
without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). We
also trained models using a smaller dataset that generated shotgun
metagenomic sequencing data from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (16). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories. Again, the performance of the models trained with the
meteagenomic data did not improve when the SCFA concentrations were
added as possible features when training the model (P\textgreater{}0.24;
Figure 2A). These data demonstrate that knowledge of the SCFA profile
from a patient's fecal sample does not improve the ability to diagnose a
colonic lesion.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (14, 37). In brief, this study used fecal
samples obtained at either a single cross-sectional time point
(n=\textbf{r
length(metai\(sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf
\)time)} - \textbf{r max(metaf\$time)} days. Our use of treatment has
been previously defined as encompassing removal of a tumor (surgery or
colonoscopy) with or without chemotherapy and radiation (37). Diagnosis
of tumor was made by colonoscopic examination and histopathological
review of biopsies obtained (14, 37). The University of Michigan
Institutional Review Board approved the study and informed consent was
obtained from all participants in accordance to the guidelines set out
by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (38). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (38) and Figure 1 here).

\textbf{16S rRNA gene sequencing.} The workflow and processing have been
previously described (37, 39, 40). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(41). Chimeras were identified and removed using UCHIME and OTUs
clustered at 97\% similarity (42). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptiClust algorithm (43).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (44) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=\textbf{r
length(metai\(sample)**) containing a total of **r length(geof_samples\)X1)}
individuals (normal n=\textbf{r filter(geof\_samples, X30 ==
``Healthy'') \%\textgreater{}\% pull(X30) \%\textgreater{}\% length()},
adenoma n=\textbf{r filter(geof\_samples, X30 == ``Adenoma'')
\%\textgreater{}\% pull(X30) \%\textgreater{}\% length()}, and carcinoma
n=\textbf{r filter(geof\_samples, X30 == ``Cancer'') \%\textgreater{}\%
pull(X30) \%\textgreater{}\% length()}) was shotgun sequenced on an
Illumina HiSeq using 125 bp paired end reads and a previously described
method (16). Briefly, the sequences were quality filtered and sequences
aligning to the human genome were removed prior to contig assembly with
MEGAHIT (45). Open Reading Frames (ORFs) were identified using Prodigal
(46), counts generated using Diamond (47), subsequent clustering into
Operational Protein Families (OPFs) used mmseq2 (48), and OPF alignment
used the KEGG database (49).

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification and regression models via the
caret package (50, 51). This was repeated on 100 different 80/20 splits
of the data to generate a reasonable range for the AUC of the model. The
reported AUCs, unless otherwise specified, are for the test sets. The
classification models were built to group normal versus adenoma, normal
versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (52). Generally, a Kruskal-Wallis rank sum
test with a Dunn's post-hoc test was used to assess differences between
individuals without colon tumors, patients with adenomas, and patients
with carcinomas. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons {[}\textbf{XXXXXX}{]}. First, we
assessed differences in SCFA concentrations measured by HPLC between
individuals with normal colons and patients with tumors (adenoma or
carcinoma). We then analyzed whether SCFA concentrations changed in
patients with an adenoma or carcinoma pre- versus post-treatment. Next,
the imputed gene counts of genes encoding enzymes involved in SCFA
synthesis was tested. Additionally, metagenomic sequencing counts for
important genes involved with SCFA production were analyzed. From here
we analyzed the number of significant positive and negative correlations
between OTU relative abundance and SCFA concentrations in individuals
without colon tumors and patients with adenomas or carcinomas using
Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs
were better able to classify individuals with and without tumors using
Random Forest models. Finally, models to classify the actual SCFA
concentration or high/low SCFA concentration based on the median of each
SCFA using 16S rRNA gene sequencing data was created using the Random
Forest algorithm. For all Random Forest models, the assessment of the
most important variables was based on the top 10 features (OTUs or
SCFAs) using the mean decrease in accuracy.

@@

\leavevmode\hypertarget{ref-Quast2012}{}%
41. \textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
J}, \textbf{Glöckner FO}. 2012. The SILVA ribosomal RNA gene database
project: Improved data processing and web-based tools. Nucleic Acids
Research \textbf{41}:D590--D596.
doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.

\leavevmode\hypertarget{ref-Edgar2011}{}%
42. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\leavevmode\hypertarget{ref-Westcott2017}{}%
43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

@@

\leavevmode\hypertarget{ref-Kanehisa2015}{}%
49. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\leavevmode\hypertarget{ref-Liaw2002}{}%
50. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

\leavevmode\hypertarget{ref-Kuhn2017}{}%
51. \textbf{Kuhn M}. 2017. Caret: Classification and regression
training.

\leavevmode\hypertarget{ref-r_citation_2017}{}%
52. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

>>>>

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models (13, 14). We repeated that analysis but
added the concentration of the individual or pooled SCFAs as possible
features to train a model. Models trained using SCFAs to classify
individuals as having adenomas or carcinomas rather than normal colons
had median areas under the receiver operator characteristic curve
(AUROC) that were significantly greater than 0.5
(P\textsubscript{adenoma}\textless{}0.001 and
P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC values
to detect the presence of adenomas or carcinomas were only 0.54 and
0.55, respectively (Figure 2A). When we trained the models with the
SCFAs concentrations and operational taxonomic unit (OTU) or genus-level
relative abundances the AUROC values were not significantly different
from the models trained without the SCFA concentrations
(P\textgreater{}0.21; Figure 2A). We also trained models using a smaller
dataset that generated shotgun metagenomic sequencing data from a subset
of our cohort (N\textsubscript{normal}=27, N\textsubscript{adenoma}=25,
and N\textsubscript{cancer}=26) (16). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories. Again, the performance of the models trained with the
meteagenomic data did not improve when the SCFA concentrations were
added as possible features when training the model (P\textgreater{}0.24;
Figure 2A). These data demonstrate that knowledge of the SCFA profile
from a patient's fecal sample does not improve the ability to diagnose a
colonic lesion.

@@

\textbf{Study design and sampling.} The overall study design and the
resulting sequence data have been previously described (14, 37). In
brief, fecal samples were obtained from 172 individuals with normal
colons, 198 individuals with colonic adenomas, and 120 individuals with
carcinomas. Of the individuals diagnosed as having adenomas or
carcinomas, a subset (N\textsubscript{adenoma}=41 and
N\textsubscript{carcinoma}=26) were sampled after treatment of the
lesion (median=255 days between sampling, IQR=233 to 334 days). Tumor
diagnosis was made by colonoscopic examination and histopathological
review of the biopsies (14). The University of Michigan Institutional
Review Board approved the studies that generated the samples and
informed consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} The measurement of acetate,
propionate, isobutyrate, and butyrate used a previously published
protocol that used High-Performance Liquid Chromatography (HPLC) (38).
Two changes were made to the protocol. First, instead of using fecal
samples suspended in DNA Genotek OmniGut tubes, we suspended frozen
fecal samples in 1 mL of PBS. Second, instead of using the average
weight of fecal sample aliquots to normalize SCFA concentrations, we
used the actual weight of the fecal samples. These methodological
changes did not affect the range of concentrations of these SCFAs
between the two studies (see Table 1 (38) and Figure 1 here).

\textbf{16S rRNA gene sequence data analysis.} Sequence data from Baxter
et al.~(14) and Sze et al.~(37) were obtained from the Sequence Read
Archive (studies SRP062005 and SRP096978) and reprocessed using using
mothur v.1.42 (39). The sequence generation workflow and processing
steps have been previously described (39, 40). In brief, the original
studies generated sequence data from V4 region of the 16S rRNA gene
using paired 250 nt reads on an Illumina MiSeq sequencer (40). The
resulting sequence data were assembled into contigs and screened for
quality aberrations including chimeras, which were removed based on the
use of UCHIME (41). Sequences were clustered into OTUs at 97\%
similarity using OptiClust (42) and assigned to the closest possible
genus using a naive Bayesian classifier with an 80\% confidence
threshold trained on the reference collection from the Ribosomal
Database Project (v.16) (43). We used PICRUSt (v.2.1.0-b) with the
recommended standard operating protocol to generate imputed metagenomes
based on the expected metabolic pathways, KEGG categories, and enzyme
commission numbers (44).

\textbf{Metagenomic DNA sequence analysis.} A subset of the samples from
the samples described by Baxter et al.~(14) were used to generate
metagenomic sequence data (N\textsubscript{normal}=27,
N\textsubscript{adenoma}=25, and N\textsubscript{cancer}=26). These data
were generated by Hannigan et al.~(16) and deposited into the Sequence
Read Archive (study SRP108915). Fecal DNA was subjected to shotgun
sequencing on an Illumina HiSeq using 125 bp paired end reads. The
archived sequences were already quality filtered and aligned to the
human genome to remove contaminating sequence data. We downloaded the
sequences and assembled them into contigs using MEGAHIT (45), which were
used to identify open reading frames (ORFs) using Prodigal (46). We
determined the abundance of each ORF by mapping the raw reads back to
the ORFs using Diamond (47). We clustered the ORFs into operational
protein families (OPFs) in which the clustered ORFs were more than 40\%
identical to each other using mmseq2 (48, 49). We also used mmseq2 to
map the ORFs to the KEGG database and clustered the ORFs according to
which category the ORFs mapped (50).

\textbf{Random Forest models.}

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.1) with
the tidyverse package (v.1.2.1). To assess differences in SCFA
concentrations between individuals normal colons and those with adenomas
or carcinomas, we used the Kruskal-Wallis rank sum test. If a test had a
P-value below 0.05, we then applied a pairwise Wilcoxon rank sum test
with a Benjamini-Hochberg correction for multiple comparisons. To assess
differences in SCFA concentrations between individuals samples before
and after treatment we used paired Wilcoxon rank sum tests to test for
significance. To compare the median AUCROC for the held out data for the
model generated using only the SCFAs, we compared the distribution of
the data to the expected median of 0.5 using the Wilcoxon rank sum test
to test whether the model performed better than would be achieved by
randomly assigning the data to each diagnosis. When we compared the
Random Forest models generated without and with SCFA data included, we
used Wilcoxon rank sum tests to determine whether the models with the
SCFA data included did better.

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at \url{https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/}.

@@

\leavevmode\hypertarget{ref-Edgar2011}{}%
41. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\leavevmode\hypertarget{ref-Westcott2017}{}%
42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\leavevmode\hypertarget{ref-Wang2007}{}%
43. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
of rRNA sequences into the new bacterial taxonomy. Applied and
Environmental Microbiology \textbf{73}:5261--5267.
doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.

@@

\leavevmode\hypertarget{ref-Schloss2008}{}%
49. \textbf{Schloss PD}, \textbf{Handelsman J}. 2008. A statistical
toolbox for metagenomics: Assessing functional diversity in microbial
communities. BMC Bioinformatics \textbf{9}.
doi:\href{https://doi.org/10.1186/1471-2105-9-34}{10.1186/1471-2105-9-34}.

\leavevmode\hypertarget{ref-Kanehisa2015}{}%
50. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.
